A fully labelled proof system for intuitionistic modal logics

### 2019-09-26 caleidoscope/caleidoscope.tex
pass through sections 1 and 2

\newcommand{\qquand}{\qquad\mbox{and}\qquad}
\newcommand{\quadcm}{\rlap{\quad,}}

@@

%\begin{frontmatter}
  \title{A properly labelled proof system for intuitionistic modal logics}
  \author{Sonia Marin \and %}
%  \todo{New university?}
%  \address{IT-Universitetet i K{\o}benhavn \\ Denmark }
%    \author{
      Marianela Morales %}
 \and
% \marianela{I think that now I have the same as Lutz, right?} No, the other order...
% \address{ LIX, \'Ecole Polytechnique  \& Inria Saclay\\ France}
%      \author{
        Lutz Stra{\ss}burger }
% \address{Inria Saclay \& LIX, \'Ecole Polytechnique \\ France}

@@

   In this paper we present a labelled sequent system for
   intuitionistic modal logics such that there is not only one, but
   two relation symbols for the labels: one for the accessibility
   relation associated with the Kripke semantics for modal logics and
   one for the preorder relation associated with the Kripke semantics
   for intuitionistic logic.
   %%
   This puts our system in close correspondence with the standard
   birelational Kripke semantics for intuitionistic modal logic. As a
   consequence it can encompass a wider range of intuitionistic modal
   logics than standard labelled systems.
   %%
   We also show an internal cut elimination proof for our system.

@@

Since their introduction in the 1980s by Gabbay~\cite{gabbay:96},
\emph{labelled proof calculi} have been widely used by proof theorist
to give sound, complete and cut-free deductive system to a broad range
of logics. Unlike so-called \emph{internal calculi}, like
hyper-sequents~\cite{avron:elc96} or nested
sequents~\cite{kashima:sl94,brunnler:aml09,poggiolesi:tmp09} or 2-sequents~\cite{masini:apal92},
labelled calculi have the advantage of being more uniform and
accommodating more logics.

Standard labelled sequent calculi attach to every formula $\fm A$ a label
$\lb x$, witten as $\labels xA$, and additionally admit so called
\emph{relational atoms} of the form $\accs xy$ where $\rel$ is a
binary relation symbol. These calculi work best for logics with
standard Kripke semantics, as in this case the relation $\rel$ is used
to encode the accessibility relation in the Kripke models, and the
frame condition corresponding to the desired logic can be directly
encoded as inference rules. Prominent examples are classical modal
logics and intuitionistic propositional logic, where, e.g., the frame
condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$), can be straightforwardly translated into the
inference rule
\begin{equation}
  \label{eq:Rtrans}
  \vlinf{\rn{trans}}{}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs xz, \Left \SEQ \Right}
\end{equation}
where $\B$ stands for a set of relational atoms, and $\Gamma$ and
$\Delta$ for multi-sets of labelled formulas.

However,
in this paper we are concerned with \emph{intuitionistic modal
  logics}, whose Kripke semantics is based on \emph{birelational}
frames, i.e., they have two binary relations instead of one: one
relation $R$ that corresponds to the accessibility relation in
ordinary Kripke frames for modal logics, and a relation $\le$ that
corresponds to the future relation in ordinary Kripke frames for
intuitionistic logic. Consequently, standard labelled systems for
these logics have certain shortcomings:
\begin{enumerate}
\item The transitivity rule in~\eqref{eq:Rtrans} corresponds to both
  versions of the $\fourax$-axiom
  \begin{equation}
    \label{eq:4ax}
    \fm{\BOX A\IMP \BOX\BOX A}
    \qquand
    \fm{\DIA\DIA A\IMP\DIA A}
    \quadcm
  \end{equation}
  as they are equivalent in classical
  modal logic. However, in intuitionistic modal logic they are not
  equivalent, and even though the logic $\ISfour$, i.e., the
  intuitionistic version of the modal logic $\Sfour$, contains both
  axioms, they can be added independently to the logic $\IK$ (the
  intuitionistic version of $\K$). However, the proof theory of these
  new logics has not been studied before, as no existing labelled (or
  label-free) proof system can handle them, even though the
  corresponding frame conditions
  \begin{equation}
    \label{eq:4frame}
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
    \qquand
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
  \end{equation}
  have already been studied
  in~\cite{plotkin:stirling:86}.\todo{check: did they get that wrong in  \cite{plotkin:stirling:86}?}
\item The correspondence between the syntax and the semantics is not
  as clean as one would expect. As only the $R$-relation (and not the
  $\le$-relation) of the frame is visible in an ordinary labelled
  sequent, we only have that a sequent $\Gamma$ is provable if and
  only if is is valid in all \emph{graph-consistent}\footnote{This
    means that every layer in the model can be lifted to any future of
    any world in that layer.}\todo{check terminology in Simpson} models, as
  already observed in by Simpson in his PhD thesis~\cite{simpson:phd}
  and considered as a very unsatisfactory situation\todo{check what he
    said exactly} (see also \cite{mar:str:tableaux17}).
\end{enumerate}
In order to address these two issues we propose here to extend the
notion of labelled sequent by also allowing relational atoms of the
form $\futs xy$. Consequently, we can easily translate the frame
conditions in~\eqref{eq:4frame} into inference rules:
\begin{equation}
  \label{eq:4rules}
  \vlinf{\rn{4^{\BOX}}}{\proviso{$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
     \B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
  \qquand
  \vlinf{\rn{4^{\DIA}}}{\proviso{$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
\end{equation}
This allows us to define cut-free deductive systems for a wide range
of logics that could not be treated before.  Furthermore, the relation
between syntax and semantics is as one would expect: A sequent is
provable in our system if and only if it is valid in all models.

Besides that, there is another pleasant observation to make about our
system. Ordinary labelled sequent systems for intuitionistic modal
logic are single-conclusion~\cite{simpson:phd}. The same is true for
the corresponding nested sequent
systems~\cite{str:fossacs13,marin:str:aiml}. It is possible to express
Maehara style multiple-conclusion systems in nested
sequents~\cite{kuz:str:maehara}, and therefore also in ordinary
labelled sequents. However, also in these systems there are rules
($\rlabrn\IMP$ and $\rlabrn\BOX$) that force a single-conclusion
premise, even though this is not the case in labelled
systems~\cite{negri:jpl2005} or nested sequents~\cite{fitting:83} for
intuitionistic logic. Our system eliminates this
discrepancy. Consequently, every rule in our system is invertible, as
we never delete information.\lutz{check, is this really the case?}

@@

This paper is organized as follows, In the next section (Section~\ref{sec:intmod}) we recall the standard syntax and semantics of intuitionistic modal logics. Then, in Section~\ref{sec:system} we present our system for the intuitionistic modal logic $\IK$. In Sections~\ref{sec:soundness} and~\ref{sec:completeness}, we show soundness and completeness of the system with cut. The cut elimination theorem, proved in Section~\ref{sec:cut-elim}, then entails soundness and completeness for the cut-free system. Finally, in Section~\ref{sec:ext} we discuss the possible extension to the system to capture other intuitionistic modal logics. \lutz{what about doing that directly in the system presentation, like Negri does it?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
\section{Intuitionistic modal logics}\label{sec:intmod}
The language of {intuitionisitic modal logic} is the one of intuitionistic propositional logic with the modal operators $\BOX$ and $\DIA$, standing most generally for \emph{necessity} and \emph{possibility}.
%
Starting with a set $\mathcal{A}$ of atomic propositions, denoted $a$, modal formulas are constructed from the grammar:
%
$$
A \coloncolonequals
a \mid A \AND A \mid A \OR A \mid \BOT \mid A \IMP A \mid \BOX A \mid \DIA A
$$
%
\todo{change color?}

@@

%The axiomatisation that is now generally accepted as intuitionistic modal logic $\IK$ was given by Plotkin and Stirling~\cite{Plotkin} and is equivalent to the one proposed by Fischer-Servi~\cite{Fischer}. It is obtained from intuitionistic propositional logic $\mathsf{IPL}$ by adding:
\begin{itemize}
	\item the \emph{necessitation rule}: if $\fm A$ is a theorem then $\fm{\BOX A}$ is also a theorem; and
	\item the following five variants of the \emph{distributivity axiom} $\rn{k}$:
	\begin{equation*}
	\label{eq:ik}%\hskip-2em
	\begin{array}[t]{r@{\;}l@{\quad}r@{\;}l@{\quad}r@{\;}l}
	\kax[1]\colon&\fm{\BOX(A\IMP B)\IMP(\BOX A\IMP\BOX B)}
	&
	\kax[3]\colon&\fm{\DIA(A\OR B)\IMP(\DIA A\OR\DIA B)}
	&
	\kax[5]\colon&\fm{\DIA\BOT\IMP\BOT}
	\\
	\kax[2]\colon&\fm{\BOX(A\IMP B)\IMP(\DIA A\IMP\DIA B)}
	&
	\kax[4]\colon&\fm{(\DIA A\IMP \BOX B)\IMP\BOX(A\IMP B)}\\%x[1ex]
	\end{array}
	\end{equation*}
\end{itemize}

The idea is that intuitionistic propositional logic does not allow the principle of \emph{Excluded Middle}, so the modalities $\BOX$ and $\DIA$ are not de Morgan duals any more, but one can choose to design the axiomatisation in order to relate them in different ways. The most basic intuitionistic modal system one can think of would be to consider only the $\BOX$ modality as regulated by the $\kax$ axiom (or as called here $\kax[1]$), which gives the system $\break \rn{IPL}$ $\rn{+}$ $\rn{nec}$ $\rn{+}$ $\kax[1]$. However this would give strictly no information on the behaviour of the $\DIA$ modality.
It seems that Fitch \cite{fitch:pm48} was the first one to propose a way to treat $\DIA$ in an intuitionistic system by considering the system  $\rn{IPL}$ $\rn{+}$ $\rn{nec}$ $\rn{+}$ $\kax[1]$ $\rn{+}$ $\kax[2]$, which is now sometimes called $\rn{CK}$ for \emph{constructive modal logic}. Wijekesera \cite{wijesekera:apal90} also considered the axiom $\kax[5]$, which states that $\DIA$ distributes over 0-ary disjunctions, but did not assume that it would always distribute over binary disjunctions; the system he proposed was therefore  $\rn{IPL}$ $\rn{+}$ $\rn{nec}$ $\rn{+}$ $\kax[1]$ $\rn{+}$ $\kax[2]$  $\rn{+}$ $\kax[5]$. The systems are well-designed for some applications for example to analyse some type systems \cite{benton:etal:jfp98} or to reason about states of a machine under partial information \cite{wijesekera:nerode:apal05}. However, they cannot satisfy the logicians as the addition of the \emph{Excluded Middle} principle to it does not yield classical modal logic $\K$, that is, it is not possible to retrieve the De Morgan duality of $\BOX$ and $\DIA$ in this case.

@@

\begin{definition}
	A \emph{bi-relational frame} $\F$ is a triple $\langle W, \rel, \le \rangle$ 
	%	of a non-empty set of worlds $W$ equipped with two binary relations $R$ and $\le$, where $R$ being the modal \emph{accessibility relation} and $\le$ a preorder (\emph{i.e.} a reflexive and transitive relation), satisfying the following conditions:
	of a set of worlds $W$ equipped with an {accessibility relation} $\rel$ and a preorder $\le$ (\emph{i.e.} a reflexive and transitive relation) satisfying:
	\begin{enumerate}
		\item[($\rn{F_1}$)] For $\lb u, \lb v, \lb{v'} \in W$, if $\accs uv$ and $\futs v{v'}$, there exists $\lb{u'}$ s.t.~$\futs u{u'}$ and $\accs{u'}{v'}$.

@@

		\item[($\rn{F_2}$)] For $\lb {u'}, \lb u, \lb v \in W$, if $\futs uv$, there exists $\lb {v'}$ s.t.~$\accs{u'}{v'}$ and $\futs v{v'}$.

@@

We write $\M, \lb w \Vdash \fm a$ if $\fm a \in V(\lb w)$, and by definition, we always have $\M, \lb w \Vdash \top$ and never that $\M, \lb w \Vdash \bot$. 
%
Then the relation is extended to all formulas by induction, following the rules for both intuitionistic and modal Kripke models:

\begin{equation*}\label{eq:kripke}
	\begin{array}{rcl}
	%\parbox{30em}{%

@@

	\end{array}
\end{equation*}
\todo{label is not appearing}

We write $\M, \lb w \not\Vdash \fm A$ if it is not the case that $\M, \lb w \Vdash \fm A$, but contrarily to the classical case, we do not have $\M, \lb w \Vdash \fm{\neg A}$ iff $\M, \lb w \Vdash \fm A$ (since $\fm{\neg A}$ is defined as $\fm{A \IMP \bot}$).

@@

\begin{definition}
	A formula $\fm A$ is \emph{satisfied} in a model $\M = \langle W, \rel, \le, V \rangle$, if for all $\lb w \in W$ we have $\M, \lb w \Vdash \fm A$.
	%
	A formula $\fm A$ is \emph{valid} in a frame $\F = \langle W, \rel, \le \rangle$, if for all valuations $V$, $\fm A$ is satisfied in $\langle W, R, \le, V \rangle$.
\end{definition}

@@

Similarly to the classical case, in the case of $\IK$, the correspondence between syntax and semantics is recovered.

\begin{theorem}[Fischer-Servi~\cite{fischer-servi:84}, Plotkin and Stirling~\cite{plotkin:stirling:86}]\label{thm:plotkin}
	A formula $\fm A$ is a theorem of $\IK$ if and only if $\fm A$ is valid in every bi-relational frame.
\end{theorem}

@@

\begin{theorem}[\cite{plotkin:stirling:86}]
	An intuitionistic modal frame $\langle W, \rel, \le \rangle$ validates $\agklmn$ if and only if the frame satisfies:

@@

In this section we present our properly labelled sequent proof system
for intuitionistic modal logics. The starting point is the notion of a
\emph{labelled formula} which is a pair $\labels xA$ of a label $\lb
x$ and a formula $\fm A$. The intended meaning is that $\fm A$ holds
at world $\lb x$. A \emph{relation atom} is either an expression
$\accs xy$ or $\futs xy$ where $\lb x$ and $\lb y$ are labels and the
intended meaning is that the corresponing relation holds between the
worlds $\lb x$ and $\lb y$. This is enough to define the notion of
\emph{(labelled) sequent}, which is a triple $\lseq\B\Left\Right$,
where $\B$ is a set of relational and pre-order atoms, and $\Left$ and $\Right$ are
multi-sets of labelled formulas, all written as list, separated by commas.

>>>>

\newcommand{\quand}{\quad\mbox{and}\quad}
\newcommand{\qquand}{\qquad\mbox{and}\qquad}
\newcommand{\quadcm}{\rlap{\quad,}}

@@

%\begin{frontmatter}
  \title{A fully labelled proof system for intuitionistic modal logics}
  \author{Sonia Marin \and %}
%  \todo{New university?}
%  \address{IT-Universitetet i K{\o}benhavn \\ Denmark }
%    \author{
      Marianela Morales %}
 \and
% \marianela{I think that now I have the same as Lutz, right?} No, the other order...
% \address{ LIX, \'Ecole Polytechnique  \& Inria Saclay\\ France}
%      \author{
        Lutz Stra{\ss}burger }
% \address{Inria Saclay \& LIX, \'Ecole Polytechnique \\ France}

@@

   In this paper we present a labelled sequent system for
   intuitionistic modal logics such that there is not only one, but
   two relation symbols for the labels: one for the accessibility
   relation associated with the Kripke semantics for normal modal logics and
   one for the preorder relation associated with the Kripke semantics
   for intuitionistic logic.
   %%
   This puts our system in close correspondence with the standard
   birelational Kripke semantics for intuitionistic modal logics. As a
   consequence it can encompass a wider range of intuitionistic modal
   logics than standard labelled systems.
   %%
   We also show an internal cut elimination proof for our system.

@@

Since their introduction in the 1980s by Gabbay~\cite{gabbay:96},
\emph{labelled proof calculi} have been widely used by proof theorist
to give sound, complete and cut-free deductive systems to a broad range
of logics. Unlike so-called \emph{internal calculi}, like
hyper-sequents~\cite{avron:elc96} or nested
sequents~\cite{kashima:sl94,brunnler:aml09,poggiolesi:tmp09}, 2-sequents~\cite{masini:apal92}, or linear nested sequents~\cite{lellmann?},
labelled calculi have the advantage of being more uniform and being able to 
accommode more logics.

Standard labelled sequent calculi attach to every formla $\fm A$ a label
$\lb x$, witten as $\labels xA$, and additionally admit so called
\emph{relational atoms} of the form $\accs xy$ where $\rel$ is a
binary relation symbol. These calculi work best for logics with
standard Kripke semantics, as in this case the relation $\rel$ is used
to encode the accessibility relation in the Kripke models, and the
frame conditions corresponding to the desired logic can be directly
encoded as inference rules. Prominent examples are classical modal
logics and intuitionistic propositional logic, where, e.g., the frame
condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$), can be straightforwardly translated into the
inference rule
\begin{equation}
  \label{eq:Rtrans}
  \vlinf{\rn{trans}}{}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs xz, \Left \SEQ \Right}
\end{equation}
where $\B$ stands for a set of relational atoms, and $\Gamma$ and
$\Delta$ for multi-sets of labelled formulas.

However,
in this paper we are concerned with \emph{intuitionistic modal
  logics}, whose Kripke semantics is based on \emph{birelational}
frames, i.e., they have two binary relations instead of one: one
relation $R$ that corresponds to the accessibility relation in
ordinary Kripke frames for modal logics, and a relation $\le$ that
corresponds to the future relation in ordinary Kripke frames for
intuitionistic logic. Consequently, standard labelled systems for
these logics have certain shortcomings:
\begin{enumerate}
\item The transitivity rule in~\eqref{eq:Rtrans} corresponds to both
  versions of the $\fourax$-axiom
  \begin{equation}
    \label{eq:4ax}
    \fm{\BOX A\IMP \BOX\BOX A}
    \qquand
    \fm{\DIA\DIA A\IMP\DIA A}
    \quadcm
  \end{equation}
  as they are equivalent in classical
  modal logic. However, in intuitionistic modal logic they are not
  equivalent, and even though the logic $\ISfour$, i.e., the
  intuitionistic version of the modal logic $\Sfour$, contains both
  axioms, they can be added independently to the logic $\IK$ (the
  intuitionistic version of $\K$). However, the proof theory of these
  new logics has not been studied before, as no existing labelled (or
  label-free) proof system can handle them, even though the
  corresponding frame conditions
  \begin{equation}
    \label{eq:4frame}
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
    \quand
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
    \;,
  \end{equation}
  respectively, have already been studied
  in~\cite{plotkin:stirling:86}.
  \item The correspondence between the syntax and the semantics is not
  as clean as one would expect. As only the $R$-relation (and not the
  $\le$-relation) of the frame is visible in an ordinary labelled
  sequent, we only have that a sequent $\Gamma$ is provable if and
  only if is is valid in all \emph{graph-consistent}\footnote{This
    means that every layer in the model can be lifted to any future of
    any world in that layer. See~\cite{simpson:phd} and~\cite{mar:str:tableaux17} for a formal definition and discussion.} for  models, as
  already observed in by Simpson in his PhD thesis~\cite{simpson:phd}
  and considered as an inelegant solution (see also \cite{mar:str:tableaux17}).
\end{enumerate}
In order to address these two issues we propose here to extend the
notion of labelled sequent by also allowing relational atoms of the
form $\futs xy$. Consequently, we can easily translate the frame
conditions in~\eqref{eq:4frame} into inference rules:
\begin{equation}
  \label{eq:4rules}
  \vlinf{\rn{4^{\BOX}}}{\proviso{$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
     \B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
  \qquand
  \vlinf{\rn{4^{\DIA}}}{\proviso{$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
\end{equation}
This allows us to define cut-free deductive systems for a wide range
of logics that could not be treated before.  Furthermore, the relation
between syntax and semantics is as one would expect: A sequent is
provable in our system if and only if it is valid in all models.

Besides that, there is another pleasant observation to make about our
system. Ordinary labelled sequent systems for intuitionistic modal
logic are single-conclusion~\cite{simpson:phd}. The same is true for
the corresponding nested sequent
systems~\cite{str:fossacs13,marin:str:aiml}. It is possible to express
Maehara style multiple-conclusion systems in nested
sequents~\cite{kuz:str:maehara}, and therefore also in ordinary
labelled sequents. However, also in these systems there are rules
($\rlabrn\IMP$ and $\rlabrn\BOX$) that force a single-conclusion
premise, even though this is not the case in labelled
systems~\cite{negri:jpl2005} or nested sequents~\cite{fitting:83} for
intuitionistic logic. Our system eliminates this
discrepancy. Consequently, every rule in our system is invertible, as
we never delete information.

@@

This paper is organized as follows, In the next section (Section~\ref{sec:intmod}) we recall the standard syntax and semantics of intuitionistic modal logics. Then, in Section~\ref{sec:system} we present our system for the intuitionistic modal logic $\IK$. In Sections~\ref{sec:soundness} and~\ref{sec:completeness}, we show soundness and completeness of the system with cut. The cut elimination theorem, proved in Section~\ref{sec:cut-elim}, then entails soundness and completeness for the cut-free system. Finally, in Section~\ref{sec:ext} we discuss the possible extension to the system to capture other intuitionistic modal logics.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
\section{Intuitionistic modal logics}\label{sec:intmod}
The language of {intuitionisitic modal logic} is the one of intuitionistic propositional logic with the modal operators $\BOX$ and $\DIA$, standing most generally for \emph{necessity} and \emph{possibility}.
%
Starting with a set $\mathcal{A}$ of atomic propositions, denoted by lower case letters $a$, $b$, $c$,\ldots, modal formulas, denoted by capital letters $A$, $B$, $C$, \ldots, are constructed from the grammar:
%
$$
A \coloncolonequals
a \mid A \AND A \mid A \OR A \mid \BOT \mid A \IMP A \mid \BOX A \mid \DIA A
$$
%
\todo{change color?}

@@

%The axiomatisation that is now generally accepted as intuitionistic modal logic $\IK$ was given by Plotkin and Stirling~\cite{Plotkin} and is equivalent to the one proposed by Fischer-Servi~\cite{Fischer}. It is obtained from intuitionistic propositional logic $\mathsf{IPL}$ by adding:
\begin{itemize}
	\item the \emph{necessitation rule}: if $\fm A$ is a theorem then $\fm{\BOX A}$ is also a theorem; and
	\item the following five variants of the \emph{distributivity axiom} $\rn{k}$:
	\begin{equation*}
	\label{eq:ik}%\hskip-2em
	\begin{array}[t]{r@{\;}l@{\qquad}r@{\;}l@{\qquad}r@{\;}l}
	\kax[1]\colon&\fm{\BOX(A\IMP B)\IMP(\BOX A\IMP\BOX B)}
	&
	\kax[3]\colon&\fm{\DIA(A\OR B)\IMP(\DIA A\OR\DIA B)}
	&
	\kax[5]\colon&\fm{\DIA\BOT\IMP\BOT}
	\\
	\kax[2]\colon&\fm{\BOX(A\IMP B)\IMP(\DIA A\IMP\DIA B)}
	&
	\kax[4]\colon&\fm{(\DIA A\IMP \BOX B)\IMP\BOX(A\IMP B)}\\%x[1ex]
	\end{array}
	\end{equation*}
\end{itemize}

The idea is that intuitionistic propositional logic does not allow the principle of \emph{Excluded Middle}, so the modalities $\BOX$ and $\DIA$ are not de Morgan duals any more, but one can choose to design the axiomatisation in order to relate them in different ways. The most basic intuitionistic modal system one can think of would be to consider only the $\BOX$ modality as regulated by the $\kax$ axiom (or as called here $\kax[1]$), which gives the system \hbox{$\rn{IPL}$ $\rn{+}$ $\rn{nec}$ $\rn{+}$ $\kax[1]$}. However this would give strictly no information on the behaviour of the $\DIA$ modality.
It seems that Fitch \cite{fitch:pm48} was the first one to propose a way to treat $\DIA$ in an intuitionistic system by considering the system  $\rn{IPL}$ $\rn{+}$ $\rn{nec}$ $\rn{+}$ $\kax[1]$ $\rn{+}$ $\kax[2]$, which is now sometimes called $\rn{CK}$ for \emph{constructive modal logic}. Wijekesera \cite{wijesekera:apal90} also considered the axiom $\kax[5]$, which states that $\DIA$ distributes over 0-ary disjunctions, but did not assume that it would always distribute over binary disjunctions; the system he proposed was therefore  $\rn{IPL}$ $\rn{+}$ $\rn{nec}$ $\rn{+}$ $\kax[1]$ $\rn{+}$ $\kax[2]$  $\rn{+}$ $\kax[5]$. The systems are well-designed for some applications for example to analyse some type systems \cite{benton:etal:jfp98} or to reason about states of a machine under partial information \cite{wijesekera:nerode:apal05}. However, they cannot satisfy the logicians as the addition of the \emph{Excluded Middle} principle to it does not yield classical modal logic $\K$, that is, it is not possible to retrieve the De Morgan duality of $\BOX$ and $\DIA$ in this case.

@@

\begin{definition}
	A \emph{bi-relational frame} $\F$ is a triple $\langle W, \rel, \le \rangle$ 
	%	of a non-empty set of worlds $W$ equipped with two binary relations $R$ and $\le$, where $R$ being the modal \emph{accessibility relation} and $\le$ a preorder (\emph{i.e.} a reflexive and transitive relation), satisfying the following conditions:
	of a set of worlds $W$ equipped with an {accessibility relation} $\rel$ and a preorder $\le$ (\emph{i.e.}\ a reflexive and transitive relation) satisfying:
	\begin{enumerate}
		\item[($\rn{F_1}$)] For all $\lb u, \lb v, \lb{v'} \in W$, if $\accs uv$ and $\futs v{v'}$, there exists $\lb{u'}$ s.t.~$\futs u{u'}$ and $\accs{u'}{v'}$.

@@

		\item[($\rn{F_2}$)] For all $\lb {u'}, \lb u, \lb v \in W$, if $\futs uv$, there exists $\lb {v'}$ s.t.~$\accs{u'}{v'}$ and $\futs v{v'}$.

@@

We write $\M, \lb w \Vdash \fm a$ if $\fm a \in V(\lb w)$, and inductively extend the $\Vdash$ relation to all formulas, following the rules for both intuitionistic and modal Kripke models:

\begin{equation}\label{eq:kripke}
	\begin{array}{rcl}
	%\parbox{30em}{%

@@

	\end{array}
\end{equation}
Observe that we never that $\M, \lb w \Vdash \fm\bot$. 
%
We write $\M, \lb w \not\Vdash \fm A$ if it is not the case that $\M, \lb w \Vdash \fm A$, but contrarily to the classical case, we do not have $\M, \lb w \Vdash \fm{\neg A}$ iff $\M, \lb w \not\Vdash \fm A$ (since $\fm{\neg A}$ is defined as $\fm{A \IMP \bot}$).

@@

\begin{definition}
	A formula $\fm A$ is \emph{satisfied} in a model $\M = \langle W, \rel, \le, V \rangle$, if for all $\lb w \in W$ we have $\M, \lb w \Vdash \fm A$.
	%
	A formula $\fm A$ is \emph{valid} in a frame $\F = \langle W, \rel, \le \rangle$, if for all valuations $V$, the formula $\fm A$ is satisfied in $\langle W, R, \le, V \rangle$.
\end{definition}

@@

Similarly to the classical case, the correspondence between syntax and semantics for $\IK$ can be stated as follows.

\begin{theorem}[\cite{fischer-servi:84,plotkin:stirling:86}]\label{thm:plotkin}
	A formula is a theorem of $\IK$ if and only if it is valid in every bi-relational frame.
\end{theorem}

@@

\lutz{maybe the foloowing goes to the ``extensions'' section. otherwise we have to define $\agklmn$ here}
\begin{theorem}[\cite{plotkin:stirling:86}]
	An intuitionistic modal frame $\langle W, \rel, \le \rangle$ validates $\agklmn$ if and only if the frame satisfies:

@@

In this section we present our fully labelled sequent proof system
for intuitionistic modal logics. The starting point is the notion of a
\emph{labelled formula} which is a pair $\labels xA$ of a label $\lb
x$ and a formula $\fm A$. The intended meaning is that $\fm A$ holds
at world $\lb x$. A \emph{relation atom} is either an expression
$\accs xy$ or $\futs xy$ where $\lb x$ and $\lb y$ are labels and the
intended meaning is that the corresponing relation holds between the
worlds $\lb x$ and $\lb y$. This is enough to define the notion of
\emph{(labelled) sequent}, which is a triple $\lseq\B\Left\Right$,
where $\B$ is a set of relational and pre-order atoms, and $\Left$ and $\Right$ are
multi-sets of labelled formulas, all written as list, separated by commas.
